### 2023-02-25 code/processing_code/1_cleaning_2020.Rmd
more cleaning

```{r}
deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
#late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)

@@

Let's start by separating `deaths` by demographic characteristic . While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together. We'll separating by sex and race/ethnicity. This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, and Hispanic persons.
```{r}
deaths_sex <- deaths %>% select(1:4, contains(c("male", "MSM")))

@@

Some rates are -9 due to missingness. Let's convert these to NA
```{r}
death_race <- deaths_race %>% 
  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
```

@@

race_pop <- us_race_pop %>%  
  pivot_longer(2:54 , names_to = "Location", values_to = "Population" ) %>%  
  pivot_wider(names_from = Label, values_from= Population)
```

@@

I need to remove all commas from the dataframe and convert the columns to numeric. 
```{r}
race_pop <- race_pop %>% 
  mutate_all(~ str_remove_all(., ",")) %>% 
  mutate(across(2:6, ~ as.numeric(.)))

```

Then, I'll calculate the counts of males and females in each location.
```{r}
total_pop <- race_pop %>% select(1:2)
pop_sex <- inner_join(total_pop, us_sex_pop)
```

@@

```{r}
kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
kid_pop <- kid_pop %>% 
  filter(TimeFrame == 2020) %>% 
  filter(`Age group` %in% c("0 to 4", "5 to 11", "12 to 14")) %>% 
  mutate(Data = as.numeric(Data))

pop_less13 <- kid_pop %>% 
  group_by(Location) %>% 
  mutate(`Age group` = ifelse(`Age group` == '12 to 14', '12', `Age group`)) %>%
  group_by(`Age group`) %>%
  mutate(`Data` = ifelse(`Age group` == '12', `Data` * 0.333, `Data`)) %>%
  group_by(Location) %>% 
  summarise(Less13 = sum(Data))

new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))

@@

df_pop <- new_pop %>% 
  group_by(Location) %>% 
  summarize(df_pop_male = sum(Male_Pop - Male_kid_pop),
            df_pop_female = sum(Female_Pop - Female_kid_pop))

```

```{r}
sex_df <- inner_join(deaths_sex, df_pop, by = join_by(State == Location))
```

```{r}
sex_df %>% mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5)
```

Next find expected rate vs observed rate

>>>>

```{r}
deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
#late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))

@@

Let's start by separating `deaths` by demographic characteristic . While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together. We'll separating by sex and race/ethnicity. This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, and Hispanic persons.
```{r}
deaths <- deaths %>% 
  filter(!grepl("DC", "State Abbreviation")) %>%
  filter(!grepl("PR", "State Abbreviation"))

deaths_sex <- deaths %>%
  select(1:4, contains("male"))

@@

Some rates are -9 due to missingness. Let's convert these to NA
```{r}
deaths_race <- deaths_race %>% 
  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
```

@@

race_pop <- us_race_pop %>%  
  pivot_longer(2:54 , names_to = "Location", values_to = "Population" ) %>%  
  pivot_wider(names_from = Label, values_from = Population)
```

@@

I need to remove all commas from the dataframe and convert the columns to numeric. 
```{r}
race_pop <- race_pop %>% 
  mutate_all(~ str_remove_all(., ",")) %>% 
  mutate(across(2:6, ~ as.numeric(.)))
```

Then, I'll calculate the counts of males and females in each location.
```{r}
total_pop <- race_pop %>%
  select(1:2)

pop_sex <- inner_join(total_pop, us_sex_pop)
```

@@

```{r}
kid_pop <- kid_pop %>% 
  filter(TimeFrame == 2020) %>% 
  filter(`Age group` %in% c("0 to 4", "5 to 11", "12 to 14")) %>% 
  mutate(Data = as.numeric(Data))

pop_less13 <- kid_pop %>% 
  group_by(Location) %>% 
  mutate(`Age group` = ifelse(`Age group` == '12 to 14', '12', `Age group`)) %>%
  group_by(`Age group`) %>%
  mutate(`Data` = ifelse(`Age group` == '12', `Data` * 0.333, `Data`)) %>%
  group_by(Location) %>% 
  summarise(Less13 = sum(Data))
```

```{r}
new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))

@@

new_pop <- new_pop %>%
  filter(!grepl("District", Location)) %>%
  filter(!grepl("United", Location)) %>% 
  filter(!grepl("Puerto Rico", Location))

df_pop <- new_pop %>% 
  group_by(Location) %>% 
  summarize(df_pop_male = sum(Male_Pop - Male_kid_pop),
            df_pop_female = sum(Female_Pop - Female_kid_pop))
```

```{r}
sex_df <- full_join(deaths_sex, df_pop, by = join_by(State == Location))
```

Let's find the HIV death rate by sex
```{r}
sex_df <- sex_df %>% 
  mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5,
         F_Rate = `Female Death Cases`/df_pop_female*1e5)
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
  drop_na(df_pop_male) %>% 
  summarize(lambda_male = sum(`Male Death Cases`)/sum(df_pop_male)) %>%
  as.numeric()

lambda_female <- sex_df %>% drop_na(`Female Death Cases`) %>% 
  drop_na(df_pop_female) %>% 
  summarize(lambda_female = sum(`Female Death Cases`)/sum(df_pop_female)) %>%
  as.numeric()

sex_df <- sex_df %>%
  mutate(Expected_Male_Cases = df_pop_male * lambda_male,
         Expected_Female_Cases = df_pop_female *  lambda_female)
```

```{r}
sex_df <- sex_df %>% 
  mutate(OE_Ratio_male = `Male Death Cases` / Expected_Male_Cases,
         OE_Ratio_female = `Female Death Cases` / Expected_Female_Cases)
```
